Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-03-11 Manuscript/manuscript.Rmd
manuscript edits

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances (bases) where a bin overlapped with genomic annotation regions (features) was calculated. In the case of no overlaps the value is set as 0.

2. **Overlap Percent (OP)**: For each genomic bin, the percent overlap between the feature width and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with the genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0.

@@

## MANY GENOMIC ANNOTATIONS WERE HIGHLY CORRELATED

When assessing the collinearity of the genomic annotations, it was found that the total number of bases that the number overlapped bases between genomic bins flanked by a TAD boundary were highly correlated between many genomic annotations (Supplementary Figure 2). That is, many genomic annotations tended to colocalize near eachother at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated with on another, yet inversely correlated with many transcription factor binding sites. The high correlations among many of the genomic annotations gives precedent for the use of an elastic-net regularization.

>>>>

1. **Overlap Counts (OC)**: For each genomic bin, the total number of instances where a genomic bin overlapped with regions assigned to genomic annotations (features) was calculated. In the case of no overlaps, the value was set to 0.

2. **Overlap Percent (OP)**: For each genomic bin, the percent of overlap between the width of the region assigned to a genomic annotation and the total width of the bin was calculated. We defined the feature width as the number of bases that overlapped with a particular genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of bases that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 10 kb, 25 kb, 50 kb, or 100 kb). Values for bins with no overlaps were set to be 0.

@@

## HIGH CORRRELATIONS IMPLY COLOCALIZATION OF MANY ANNOTATIONS AROUND TAD BOUNDARIES

We assessed the collinearity among the genomic annotations used for the feature space of subsequent models by isolating the genomic bins that overlapped with flanked TAD boundaries. We then vectorized the number of instances when a region defined by a genomic annotation overlapped with the isolated bins. It was found that the correlations among these vectors were very high for many genomic annotations. The correlations were visualized using correlograms for each data resolution and cell line (Supplementary Figure 2). The high correlations imply that many genomic annotations tended to colocalize near eachother at the boarders of domains. This was present across both resolution and cell line. Specifically, we found that the CTCF, Rad21, Smc3ab, and Znf143 transcription factor binding sites were highly colocalized near TAD boundaries. Likewise, many histone modifications also tended to appear near TAD boundaries in conjunction with one another including H3K4me1, H3K4me2, H3K4me3, H3K9ac, and H3K27ac. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly correlated with on another, yet inversely correlated with many transcription factor binding sites.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. The regularization was performed after implementing each of the resampling techniques on all data resolutions within each cell line. Additionally, a unique predictor-type was included separately as the set of features. It was found that the elastic-net regularized the coefficients of far more distance-type predictors to 0 compared to datasets with overlap count and percent predictor types, regardless of cell line or data resolution (Supplementary Figure 3). Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.
